---
title: "PA1_template.Rmd"
author: "Marcus A. Streips"
date: "January 6, 2015"
output: html_document
---
###Loading and Preprocessing the Data###

After setting the working directorty using the ```setwd()``` command, the following libraries were loaded:  


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(plyr)
library(timeDate)
library(lattice)
```
The data was loaded from the "activity.csv" file using the `read.csv()` command and the date variable was converted from factor to date using `as.Date()`. 

```{r}
data<-read.csv("activity.csv")
data$date <- as.Date(data$date)
```

###What is the mean total number of steps taken per day?###
For this part of the report, we ignore the missing values in the dataset.  


```{r}
hist(data$steps, main="Histogram of Total Number\n of Steps Taken per Day", 
     xlab="Steps Taken per Day")
```

The `summary()` command is used calculate the median and mean **total** number of steps taken per day (0.00 and 37.38 respectively). 
```{r}
summary(data$steps)
```

###What is the average daily activity pattern?###
using the `aggregate()` command, we calculate the mean of steps for each interval.
```{r}
data_avg <- aggregate(steps~interval, data=data, FUN= "mean")
plot(data_avg$interval, data_avg$steps, type="l", 
     main="Average Daily Activity Pattern\nby 5-Minute Intervals", 
     xlab="5 Minute Invervals",
     ylab="Average Number of Steps Across All Days")
```

We use this funky looking command to determine which 5-minute interval average contains the maximum number of steps (835). 
```{r}
data_avg[which.max(data_avg$steps),]
```

###Imputing Missing Values###
The total number of missing values calculated by variable (2304):
```{r}
summary(is.na(data))
```
The mean for each 5-minute interval is used to fill missing values using the `ddply()` command and arranged on "date" and "interval".
```{r}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
data_new2 <- ddply(data, ~ interval, transform, steps = impute.mean(steps))
data_new3 <- arrange(data_new2, date, interval)
head(data_new3, 5)
```

As we see from the histogram of the newly created dataset and from the data summary that the data has changed reflecting more steps taken in total (3rd Qu. jumped from 12.0 to 27.0 steps), but the mean and median for the overall dataset remained the same.

```{r}
hist(data_new3$steps, main="Histogram of Total Number\n of Steps Taken per Day", 
     xlab="Steps Taken per Day")
```

```{r}
summary(data_new3$steps)
```

###Differences in Activity between Weekdays and Weekends###
There is likely a more elegant way of processing the data, but I decided to separate the weekday intervals from the weekend intervals, process each in turn, and then combine them again to plot the comparison on lattice. 

**Separating Weekdays from Weekends**  
  

Using the `mutate()` command in combination with `isWeekday()` from the `tableDate` library I separate each interval into two groups:
```{r}
data_new5 <- mutate(data_new3, weekday=isWeekday(date))
data_wkdy <- filter(data_new5, weekday==TRUE)
data_wknd <- filter(data_new5, weekday==FALSE)
head(data_wkdy, 5)
head(data_wknd, 5)
```

**Calculating Means Over Intervals for Each Group**  
  
The `aggregate()` command is used, as before, to determine the average over the relevant time period for each interval. 

```{r}
data_wk_avg <- aggregate(steps~interval, data=data_wkdy, FUN= "mean")
data_wknd_avg <- aggregate(steps~interval, data=data_wknd, FUN= "mean")
```

**Final Processing for Lattice Plot**  
  
My data is ready to be plotted and compared, but, first I will need to label each interval with the appropriate designation "weekday/weekend" and transform the designations into a 2-level factor variable "daytype"
```{r}
data_wk_avg2 <- mutate(data_wk_avg, daytype="weekday")
data_wknd_avg2 <-mutate(data_wknd_avg, daytype="weekend")
mergedData <- rbind(data_wknd_avg2, data_wk_avg2)
mergedData <- transform(mergedData, daytype = factor(daytype))
head(mergedData, 5)
```

We are now ready to plot using the lattice system (compare with assignment Github [Readme](https://github.com/mastreips/RepData_PeerAssessment1))

```{r}
xyplot(steps ~ interval | daytype, data= mergedData, layout = c(1,2), type="l",
       main="Weekend and Weekday Activity Patterns",
       xlab = "5-Minute Intervals",
       ylab= "Average Number of Steps Across All Days")
```