---
title: "PA1_template.Rmd"
author: "Marcus A. Streips"
date: "January 6, 2015"
output: html_document
---
###Loading and Preprocessing the Data###

After setting the working directorty using the ```setwd()``` command, the following libraries were loaded:  


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(plyr)
library(timeDate)
library(lattice)
```
The data was loaded from the "activity.csv" file using the `read.csv()` command and the date variable was converted from factor to date using `as.Date()`. 

```{r}
data_tmp<-read.csv("activity.csv")
data_tmp$date <- as.Date(data_tmp$date)
```

We then must find the total number of steps taken by day intervals.

```{r}
data <- aggregate(data_tmp$steps , 
               list(date = cut(as.Date(data_tmp$date), 
                              breaks="1 day")), FUN=sum)
```

###What is the mean total number of steps taken per day?###
For this part of the report, we ignore the missing values in the dataset.  


```{r}
hist(data$x, main="Histogram of Total Number\n of Steps Taken per Day", 
     xlab="Steps Taken per Day",
     breaks = 20,
     xlim= c(0,25000),
     ylim= c(0,20))
```

The `summary()` command is used calculate the median and mean **total** number of steps taken per day (10,760 and 10,770 respectively). 
```{r}
summary(data$x)
```

###What is the average daily activity pattern?###
using the `aggregate()` command, we calculate the mean of steps for each interval.
```{r}
data_avg <- aggregate(steps~interval, data=data_tmp, FUN= "mean")
plot(data_avg$interval, data_avg$steps, type="l", 
     main="Average Daily Activity Pattern\nby 5-Minute Intervals", 
     xlab="5 Minute Invervals",
     ylab="Average Number of Steps Across All Days")
```

We use this funky looking command to determine which 5-minute interval average contains the maximum number of steps (835). 
```{r}
data_avg[which.max(data_avg$steps),]
```

###Imputing Missing Values###
The total number of missing values calculated by variable (2304):
```{r}
summary(is.na(data_tmp))
```
The mean for each 5-minute interval is used to fill missing values using the `ddply()` command and arranged on "date" and "interval".
```{r}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
data_new <- ddply(data_tmp, ~ interval, transform, steps = impute.mean(steps))
data_new2 <- arrange(data_new, date, interval)
head(data_new2, 5)
```
As before, we must `aggregate()` the values of the new data set by day.
```{r}
data_new3 <- aggregate(data_new2$steps , 
                  list(date = cut(as.Date(data_new2$date), 
                                  breaks="1 day")), FUN=sum)
```
As we see from the histogram of the new data set, the frequency of the mean total daily steps is increased - since the mean of the 5-min intervals summed on a daily basis will be the mean of total daily steps.

```{r}
hist(data_new3$x,
     main="Histogram of Total Number\n of Steps Taken per Day", 
     xlab="Steps Taken per Day",
     breaks = 20,
     xlim= c(0,25000),
     ylim= c(0,20)
     )
```
  
  
Because of the imputed values, the mean stays the same, but the median changes (now both mean and median are 10,770).
```{r}
summary(data_new3$x)
```

###Differences in Activity between Weekdays and Weekends###
There is likely a more elegant way of processing the data, but I decided to separate the weekday intervals from the weekend intervals, process each in turn, and then combine them again to plot the comparison on lattice. 

**Separating Weekdays from Weekends**  
  

Using the `mutate()` command in combination with `isWeekday()` from the `tableDate` library on the new dataset, I separate each interval into two groups:
```{r}
data_new5 <- mutate(data_new2, weekday=isWeekday(date))
data_wkdy <- filter(data_new5, weekday==TRUE)
data_wknd <- filter(data_new5, weekday==FALSE)
head(data_wkdy, 5)
head(data_wknd, 5)
```

**Calculating Means Over Intervals for Each Group**  
  
The `aggregate()` command is used, as before, to determine the average over the relevant time period for each interval. 

```{r}
data_wk_avg <- aggregate(steps~interval, data=data_wkdy, FUN= "mean")
data_wknd_avg <- aggregate(steps~interval, data=data_wknd, FUN= "mean")
```

**Final Processing for Lattice Plot**  
  
My data is ready to be plotted and compared, but, first I will need to label each interval with the appropriate designation "weekday/weekend" and transform the designations into a 2-level factor variable "daytype"
```{r}
data_wk_avg2 <- mutate(data_wk_avg, daytype="weekday")
data_wknd_avg2 <-mutate(data_wknd_avg, daytype="weekend")
mergedData <- rbind(data_wknd_avg2, data_wk_avg2)
mergedData <- transform(mergedData, daytype = factor(daytype))
head(mergedData, 5)
```

We are now ready to plot using the lattice system (compare with assignment Github [Readme](https://github.com/mastreips/RepData_PeerAssessment1))

```{r}
xyplot(steps ~ interval | daytype, data= mergedData, layout = c(1,2), type="l",
       main="Weekend and Weekday Activity Patterns",
       xlab = "5-Minute Intervals",
       ylab= "Average Number of Steps Across All Days")
```